### 2017-01-31 real/draft.tex
[real] Fix several bugs in moments calculation
[draft] Discuss asymptotic plot in more detail

\newcommand{\refeq}[1]{Eq.~(\ref{#1})}
\newcommand{\reffig}[1]{Fig.~\ref{fig:#1}}
\newcommand{\refsec}[1]{Sec.~\ref{sec:#1}}

\DeclareMathOperator{\Expect}{\mathbb{E}}
\newcommand{\expect}[1]{\Expect\left[#1\right]}
\newcommand{\expectest}[1]{\widehat{\Expect\left[#1\right]}}
\DeclareMathOperator{\GammaDist}{Gamma}
\DeclareMathOperator{\GaussianDist}{\mathcal{N}}
\DeclareMathOperator{\InvGammaDist}{InvGamma}
\newcommand{\Kalpha}{{K_\alpha}}
\newcommand{\Kbeta}{{K_\beta}}
\newcommand{\npack}{{n_p}}
\newcommand{\lmax}{\ell_{\rm max}}
\newcommand{\Lmax}{{L_{\rm max}}}
\newcommand{\Lumtot}{Q}
\newcommand{\lumtot}{q}
\newcommand{\Lum}{L}
\newcommand{\lum}{\ell}
% \newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential
\newcommand{\rmdx}[1]{\dd{#1}} % differential
\newcommand{\firstDeriv}[1]{\frac{\partial}{\partial #1}}
\newcommand{\secDeriv}[1]{\frac{\partial^2}{\partial #1^2}} % second partial derivative
\newcommand{\secPartial}[2]{\frac{\partial^2}{\partial #1 \, \partial #2}} % second partial derivative
\newcommand{\tardis}{TARDIS}
\DeclareMathOperator{\Variance}{\mathbb{V}}
\newcommand{\variance}[1]{\Variance\left[#1\right]}

@@

We return to the general formulation of \refeq{sll} and re-arrange its
factors into
\begin{align}
  \lleq{asc}
  p(Q\cond n,\bml)
  &= \int \rmdx{\lambda}\; \rmdx{\bmphi}\;
    p(\lambda\cond n)\;
    p(\bmphi\cond n,\bml)\;
    p(Q\cond \lambda,\bmphi) \\
  \lleq{asd}
  p(Q\cond \lambda,\bmphi)
  &= \sum_N
    p(N\cond \lambda)\;
    \int \rmdx{\bmL}\;
    p(Q\cond N,\bmL)\;
    p(\bmL\cond N,\bmphi)
\end{align}
with $p(\bmL\cond N,\bmphi) = \prod_{j=1}^N p(L_j\cond\bmphi)$ as
before. Again all quantities are related to a single bin in frequency.
Writing the moments and variance of $L$ as\footnote{The $k$-th moment
  of variable $x$ is defined as the expectation value $\mu_k =
  \expect{x^k} = \int \rmdx{x}\,x^k\,p(x),\; k = 0,1,2,\ldots$.} %
\begin{align}
  \lleq{ram}
  \mu =  \mu_1 &\equiv \expect{\Lum}, \qquad
  \mu_2 \equiv \expect{\Lum^2}, \\
  \sigma^2 &\equiv \variance{\Lum}
  = \expect{\Lum^2} - \expect{\Lum}^2,
\end{align}
it is easy to show that, for the Poissonian compound process
(\ref{asd}), the first moment and variance of $Q$ are related to those
of $L$ by
%%%%%%%%%%%%%%%%
\fredmargin{citation} %
%%%%%%%%%%%%%%%
\begin{align}
  \lleq{raa}
  \expect{Q} = \lambda \mu, \qquad
  \variance{Q} = \lambda \mu_2 = \lambda (\mu^2 + \sigma^2) \,.
\end{align}
for any probability $p(L\cond\bmphi)$.
%
Higher-order quantities such as the skewness and kurtosis
are suppressed by powers of $\lambda^{-1/2}$.
%%%%%%%%%%%
\hans{The variance is the second so-called \textit{cumulant}. The
  relation for all orders is that the $k$-th order cumulant
  $\kappa_q(Q)$ is equal to $\lambda$ times the $k$-th order moment of
  $L$, $\mu_k$. The skewness is therefore
  $\kappa_3(Q)/\kappa_2(Q)^{3/2} = \lambda \mu_3/(\lambda \mu_2)^{3/2}
  = \lambda^{-1/2} \mu_3/\mu_2^{3/2}$. The magnitude of the correction
  to (rab) due to $\kappa_3$ and higher orders is unclear.}
%%%%%%%%%%%
From Theorem 4.3.1 of Ref.~\cite{bening2002generalized}, we conclude
that the asymptotic distribution of $\Lumtot$ when
$\lambda \to \infty$, given the first and second moment of $\Lum$, is
a Gaussian or normal distribution,
\begin{align}
  \lleq{rab}
  P(\Lumtot \cond \lambda,\bmphi)
  &= \GaussianDist \left( \Lumtot \cond \expect{Q}, \variance{Q} \right) %
    = \GaussianDist \left( \Lumtot \cond \lambda \mu_1, \lambda (\mu^2 {+} \sigma^2) \right).
\end{align}
This drastic simplification eliminates both the sum over $N$ and the
integral over $\bmL$ in \refeq{asd}.

@@

\fred{In \reffig{asymptotic}, we contrast the asymptotic expression
  with the ``exact'' expression \refeq{spk} where the integrals over
  $\bmphi$ are done numerically to a requested relative accuracy of
  $10^{-5}$ through the Cubature
  package\cite{cubature,genz1980remarks,berntsen1991adaptive}. We
  computed $P(\Lumtot \cond n, \bml)$ on a grid of $\Lumtot$ values
  and checked if the probability integrates to one using Simpson's
  rule. With the exact expression, we obtain 0.997 ($n=10$) and 0.996
  ($n=80$), and 0.698 and 0.953 for the asymptotic expression. For
  $n=10$, the large discrepancy arises mostly because the constraint
  $P(Q \leq 0 \cond n, \bml) = 0$ is not respected in \refeq{ral}
  whereas it is satisfied by
  \refeq{spk}. Following~\cite{tierney_accurate_1986}, we renormalized
  the results to exactly integrate to one for ease of comparison in
  \reffig{asymptotic}. Note that the lack of normalization may not be
  relevant; e.g. if \refeq{ral} is used as the likelihood for a single
  bin in Bayesian parameter inference.

  We verified that for $n>80$, the two methods agree even better than
  shown there and that the normalization of the asymptotic result
  approaches one. }

@@

While the integral in \refeq{ral} cannot be integrated analytically,
the integrand is unimodal and strongly peaked, so that the Laplace or
saddle-point approximation~\cite[Ch. 27]{mackay2003information} can be
applied. If only the value $p(\Lumtot{=}\lumtot \cond n,m_1,m_2)$ at the
total data luminosity $q$ is sought, then one can approximate it
directly. In addition, we can compute the moments of $\Lumtot$ through
the \emph{moment generating function} \cite{stuart1994kendall}
\begin{align}
  \lleq{rap}
  M(t \cond \bmy) = \int_0^{\infty} \rmdx{\Lumtot} p(\Lumtot \cond \bmy) e^{-t \Lumtot} \,
\end{align}
from which the moments are accessible by differentiation
\begin{align}
  \lleq{raq}
  \expect{\Lumtot^k} = (-1)^k \eval{\dv[k]{M(t \cond \bmy)}{t}}_{t=0}
\end{align}
Assuming all involved integrals are absolutely convergent, we can
exchange the order of integration and find
\begin{align}
  \lleq{rar}
  M(t \cond \bmy) &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond \bmy) p( \mu, \sigma^2 \cond \bmy)  f(t \cond \lambda, \mu, \sigma^2),
  \intertext{where }
  f\left(t \cond  \lambda, \mu, \sigma^2\right) &\equiv \int \rmdx{\Lumtot} \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda \sigma^2 \right) e^{-t \Lumtot}= \frac{1}{2} e^{t \lambda (t \sigma^2/2 - \mu)} \left[ 1 + \erf\qty( \frac{\sqrt{\lambda}(\mu - t \sigma^2)}{\sqrt{2 \sigma^2}}) \right].
\end{align}
This implies that the mean and variance of $\Lumtot$ require an integral over the three parameters
\begin{align}
  \lleq{ras}
  \expect{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond n) p( \mu, \sigma^2 \cond \bmy) \qty(-\eval{\dv{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0})\\
  \variance{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond n) p( \mu, \sigma^2 \cond \bmy) \qty(\eval{\dv[2]{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0}) - \expect{\Lumtot}^2
\end{align}
We implemented the above formulae in the julia language~\cite{julia14}
and used Newton's method from the \texttt{Optim.jl} package in
combination with automatic
differentiation~\cite{RevelsLubinPapamarkou2016} to optimize the
respective integrand and compute the Hessian at the mode. As initial
values, we slightly offset the sample estimates to avoid issues if the
optimizer cannot improve on the initial guess and found convergence
within $\order{10}$ steps:
\begin{align}
  \lleq{rat}
  \qty(\lambda, \mu, \sigma^2) = 1.0005 (n, \langle\lum\rangle, \langle\lum^2\rangle{-}\langle\lum\rangle^2).
\end{align}
\fred{Did we define expectation value and variance? Should be in notation or intro section. Hans: see footnote 2.} %
\fred{At least for us, I should do a comparison plot to evaluate $p(\Lumtot \cond \bmy)$ with both methods } %
\fred{We need a plot, the one to compare predictions and 250 real replicas or 100 virtual replicas?} %
\fred{Advantage of this approach: no need to store all packets anymore, sum and sum of squares can be accumulated online $\Rightarrow$ reduced computation and storage requirements}

>>>>

\newcommand{\refeq}[1]{Eq.~(\ref{#1})}
\newcommand{\reffig}[1]{Fig.~\ref{fig:#1}}
\newcommand{\refsec}[1]{Sec.~\ref{sec:#1}}
\DeclareMathOperator{\erfc}{erfc}
\DeclareMathOperator{\Expect}{\mathbb{E}}
\newcommand{\expect}[1]{\Expect\left[#1\right]}
\newcommand{\expectest}[1]{\widehat{\Expect\left[#1\right]}}
\DeclareMathOperator{\GammaDist}{Gamma}
\DeclareMathOperator{\GaussianDist}{\mathcal{N}}
\DeclareMathOperator{\InvGammaDist}{InvGamma}
\newcommand{\Kalpha}{{K_\alpha}}
\newcommand{\Kbeta}{{K_\beta}}
\newcommand{\npack}{{n_p}}
\newcommand{\lmax}{\ell_{\rm max}}
\newcommand{\Lmax}{{L_{\rm max}}}
\newcommand{\Lumtot}{Q}
\newcommand{\lumtot}{q}
\newcommand{\Lum}{L}
\newcommand{\lum}{\ell}
% \newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential
\newcommand{\rmdx}[1]{\dd{#1}} % differential
\newcommand{\firstDeriv}[1]{\frac{\partial}{\partial #1}}
\newcommand{\secDeriv}[1]{\frac{\partial^2}{\partial #1^2}} % second partial derivative
\newcommand{\secPartial}[2]{\frac{\partial^2}{\partial #1 \, \partial #2}} % second partial derivative
\newcommand{\tardis}{TARDIS}
\DeclareMathOperator{\Variance}{\mathbb{V}}
\newcommand{\variance}[1]{\Variance\left[#1\right]}

@@

We return to the general formulation of \refeq{sll} and re-arrange its
factors into
\begin{align}
  \lleq{asc}
  p(Q\cond n,\bml)
  &= \int \rmdx{\lambda}\; \rmdx{\bmphi}\;
    p(\lambda\cond n)\;
    p(\bmphi\cond n,\bml)\;
    p(Q\cond \lambda,\bmphi) \\
  \lleq{asd}
  p(Q\cond \lambda,\bmphi)
  &= \sum_N
    p(N\cond \lambda)\;
    \int \rmdx{\bmL}\;
    p(Q\cond N,\bmL)\;
    p(\bmL\cond N,\bmphi)
\end{align}
with $p(\bmL\cond N,\bmphi) = \prod_{j=1}^N p(L_j\cond\bmphi)$ as
before. Again all quantities are related to a single bin in frequency.
Writing the moments and variance of $L$ as\footnote{The $k$-th moment
  of variable $x$ is defined as the expectation value $\mu_k =
  \expect{x^k} = \int \rmdx{x}\,x^k\,p(x),\; k = 0,1,2,\ldots$.} %
\begin{align}
  \lleq{ram}
  \mu =  \mu_1 &\equiv \expect{\Lum}, \qquad
  \mu_2 \equiv \expect{\Lum^2}, \\
  \sigma^2 &\equiv \variance{\Lum}
  = \expect{\Lum^2} - \expect{\Lum}^2,
\end{align}
it is easy to show that, for the Poissonian compound process
(\ref{asd}), the first moment and variance of $Q$ are related to those
of $L$ by
%%%%%%%%%%%%%%%%
\fredmargin{citation} %
%%%%%%%%%%%%%%%
\begin{align}
  \lleq{raa}
  \expect{Q} = \lambda \mu, \qquad
  \variance{Q} = \lambda \mu_2 = \lambda (\mu^2 + \sigma^2) \,.
\end{align}
for any probability $p(L\cond\bmphi)$.
%
Higher-order quantities such as the skewness and kurtosis
are suppressed by powers of $\lambda^{-1/2}$.
%%%%%%%%%%%
\hans{The variance is the second so-called \textit{cumulant}. The
  relation for all orders is that the $k$-th order cumulant
  $\kappa_q(Q)$ is equal to $\lambda$ times the $k$-th order moment of
  $L$, $\mu_k$. The skewness is therefore
  $\kappa_3(Q)/\kappa_2(Q)^{3/2} = \lambda \mu_3/(\lambda \mu_2)^{3/2}
  = \lambda^{-1/2} \mu_3/\mu_2^{3/2}$. The magnitude of the correction
  to (rab) due to $\kappa_3$ and higher orders is unclear.}
%%%%%%%%%%%
From Theorem 4.3.1 of Ref.~\cite{bening2002generalized}, we conclude
that the asymptotic distribution of $\Lumtot$ when
$\lambda \to \infty$, given the first and second moment of $\Lum$, is
a Gaussian or normal distribution,
\begin{align}
  \lleq{rab}
  P(\Lumtot \cond \lambda,\bmphi)
  &= \GaussianDist \left( \Lumtot \cond \expect{Q}, \variance{Q} \right) %
    = \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\mu^2 {+} \sigma^2) \right).
\end{align}
This drastic simplification eliminates both the sum over $N$ and the
integral over $\bmL$ in \refeq{asd}.

@@

In \reffig{asymptotic}, we contrast the asymptotic expression with the
``exact'' expression \refeq{spk} where the integrals over $\bmphi$ are
done numerically to a requested relative accuracy of $10^{-5}$ through
the Cubature
package\cite{cubature,genz1980remarks,berntsen1991adaptive}. We
computed $P(\Lumtot \cond n, \bml)$ on a grid of $\Lumtot$ values and
checked if the probability integrates to one using Simpson's
rule. With the exact expression, we obtain 0.997 ($n=10$) and 0.996
($n=80$), and 0.698 and 0.953 for the asymptotic expression. For
$n=10$, the large discrepancy arises mostly because the positivity
constraint $P(Q \leq 0 \cond n, \bml) = 0$ is not respected in
\refeq{ral} since the Gaussian approximation \refeq{rae} allows
negative $\lambda$ and $\Lumtot$, whereas the results \refeq{spk}
based on the $\GammaDist$ model imply $\lambda \geq 0$ and $Q \geq 0$.

For that reason, we discourage the asymptotic approximation for small
$n$ but we note that in general, the lack of numerical normalization
may not be a concern. For example, if \refeq{ral} is used as the
likelihood for a single bin in Bayesian parameter inference, the
normalization or evidence is
irrelevant. Following~\cite{tierney_accurate_1986}, we renormalized
the results to integrate to one for ease of comparison in
\reffig{asymptotic}.  We verified that for $n>80$, the two methods
agree even better than shown there and that the normalization of the
asymptotic result approaches one.

The main advantage of the approximation is computational efficiency. A
single evaluation of $P(Q \cond n, \bml)$ takes only about 100 calls
to the integrand of \refeq{ral} regardless of the value of $n$. In
contrast, a the numerical integration of a single term for fixed $N$
in \refeq{spk} requires 1000s of function evaluations, and the number
of terms grows like $\sqrt{n}$ for a given accuracy of
$P(Q \cond n, \bml)$.

@@

While the integral in \refeq{ral} cannot be integrated analytically,
the integrand is unimodal and strongly peaked, so that the Laplace or
saddle-point approximation~\cite[Ch. 27]{mackay2003information} can be
applied. If only the value $p(\Lumtot{=}\lumtot \cond n,m_1,m_2)$ at the
total data luminosity $q$ is sought, then one can approximate it
directly. In addition, we can compute the moments of $\Lumtot$ through
the \emph{moment generating function} \cite{stuart1994kendall}
\begin{align}
  \lleq{rap}
  M(t \cond \bmy) = \int_0^{\infty} \rmdx{\Lumtot} p(\Lumtot \cond \bmy) e^{-t \Lumtot} \,
\end{align}
from which the moments are accessible by differentiation
\begin{align}
  \lleq{raq}
  \expect{\Lumtot^k} = (-1)^k \eval{\dv[k]{M(t \cond \bmy)}{t}}_{t=0}
\end{align}
Assuming all involved integrals are absolutely convergent, we can
exchange the order of integration and find
\begin{align}
  \lleq{rar}
  M(t \cond \bmy) &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond \bmy) p( \mu, \sigma^2 \cond \bmy)  f(t \cond \lambda, \mu, \sigma^2),
  \intertext{where }
  f\left(t \cond  \lambda, \mu, \sigma^2\right) &\equiv \int \rmdx{\Lumtot} \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\mu^2 + \sigma^2) \right) e^{-t \Lumtot}\\
  &= \frac{1}{2} \exp \qty(t/2 \lambda (t (\mu^2 + \sigma^2)- 2\mu)) \erfc \qty( \frac{\sqrt{\lambda}(t \qty(\mu^2 + \sigma^2) - \mu)}{\sqrt{2 \qty(\mu^2 + \sigma^2)}}).
\end{align}
This implies that the mean and variance of $\Lumtot$ require an integral over the three parameters
\begin{align}
  \lleq{ras}
  \expect{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond n) p( \mu, \sigma^2 \cond \bmy) \qty(-\eval{\dv{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0})\\
  \variance{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond n) p( \mu, \sigma^2 \cond \bmy) \qty(\eval{\dv[2]{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0}) - \expect{\Lumtot}^2
\end{align}
We implemented the above formulae in the julia language~\cite{julia14}
and used Newton's method from the \texttt{Optim.jl} package in
combination with automatic
differentiation~\cite{RevelsLubinPapamarkou2016} to optimize the
respective integrand and compute the Hessian at the mode. As initial
values, we slightly offset the sample estimates to avoid issues if the
optimizer cannot improve on the initial guess and found convergence
within $\order{10}$ steps:
\begin{align}
  \lleq{rat}
  \qty(\lambda, \mu, \sigma^2) = 1.0005 (n, \langle\lum\rangle, \langle\lum^2\rangle{-}\langle\lum\rangle^2).
\end{align}
\fred{At least for us, I should do a comparison plot to evaluate $p(\Lumtot \cond \bmy)$ with both methods } %
\fred{We need a plot, the one to compare predictions and 250 real replicas or 100 virtual replicas?} %
\fred{Advantage of this approach: no need to store all packets anymore, sum and sum of squares can be accumulated online $\Rightarrow$ reduced computation and storage requirements}
